---
title: "Assigment: Trend Surfaces"
date: "`r Sys.Date()`"
author: "Yunxuan Pei"
output:
   html_document:
     toc: TRUE
     toc_float: TRUE
     theme: "united"
     code_folding: show
     number_section: TRUE
     fig_height: 6
     fig_width: 8
     fig_align: 'center'
---

# Dataset
***

Task 1: Load the "Northeast Climate" R Dataset `ne_stations.RData`, with the `load` function. Use the `verbose=TRUE` optional argument to display the names of the objects that are loaded.

```{r Task 1}
load("ne_stations.RData", verbose = TRUE)
```

<br>

Q1: What are the data types of these objects? For those that are spatial, what are their coördinate reference systems (CRS)?

Note: you will have to load the `sp` package to see the CRS.

A:

```{r Q1(1), eval=FALSE}
install.packages("sp")
```

```{r Q1(2)}
library("sp")
class(ne)
proj4string(ne)
class(ne.m)
proj4string(ne.m)
class(ne.df)
class(state.ne)
proj4string(state.ne)
class(state.ne.m)
proj4string(state.ne.m)
class(ne.crs)
```

We will work mostly with object `ne.df`, a dataframe version of the climate stations.

<br>

Q2: For how many climate stations are there records? How many states of the USA are represented?

A: 

```{r Q2}
length(ne.df$STATION_ID)
levels(ne.df$STATE)
```
- There are 305 climate stations in the records. 
- There are 4 states (NJ, NY, PA, VT) are represented.

<br>

Task 2: Display the names of the variables in this dataset.

```{r Task 2}
names(ne.df)
```

<br>

Q3: What are the attributes in the `ne.df` data frame?

A: 

-  Attributes in the `ne.df` data frame are "STATION_ID" "STATE" "STATION_NA" "LATITUDE_D" "LONGITUDE_" "ELEVATION_" "OID_" "COOP_ID" "STATE_1" "STN_NAME" "LAT_DD" "LONG_DD" "ELEV_FT" "JAN_GDD50" "FEB_GDD50" "MAR_GDD50" "APR_GDD50" "MAY_GDD50" "JUN_GDD50" "JUL_GDD50" "AUG_GDD50" "SEP_GDD50" "OCT_GDD50" "NOV_GDD50" "DEC_GDD50" "ANN_GDD50" "E" "N". 

The response variable here is `ANN_GDD50` (annual growing-degree days), the coördinates are in fields `E` and `N`, the elevation of the station is in field `ELEVATION_`.

<br>

# First-order trend surface
***

Task 3: Compute a first-order trend surface model using the two coordinates (East and North) as the predictors, and the growing degree days as the dependent (response) variable, using Ordinary Least Squares linear regression.

```{r Task 3(1)}
model_t3 <- lm(ANN_GDD50 ~ E + N, data = ne.df)
```

Display its summary and the linear model diagnostic plots "residuals vs. fitted" and "Normal Q-Q".

```{r Task 3(2)}
summary(model_t3)
par(mfrow = c(1,2))
plot(model_t3, which = 1:2)
```

<br>

Q4: What proportion of the variation in GDD50 is explained by the first-order trend?

A: 

- The proportion of variability explained is 0.58. 

<br>

Q5: Which of the two coördinates most affects the GDD50?

A: 

- For each km E it increase by 7.483e-04 m (0.0007483 m); for each km N it decrease by -2.818e-03 m (-0.002818 m). They both are not that important to the predictants. Compared to E, N is kind of more affected to the GDD50.

<br>

Q6: Is there a relation between the fitted values and residuals? If so, what?

A: 

- There is a relation between residuals and fitted values: residuals at both extremes are positive (under-predictions); in the mid-range most residuals are negative (over-predictions). Mean residual is not 0 through the range of fitted values.

<br>

Q7: Are the residuals normally-distributed? If not, how so? Are there any especially large residuals?

A: 

- Extreme residuals are not from a normal distribution. There are some extreme value influence the residuals distribution.

<br>

Optional Task: Identify the most poorly-modelled station.

Optional Q: Where is this? Why might it be so poorly modelled?

Optional A: 

Task 4: Display a post-plot of the residuals, with the positive residuals (under-predictions) shown in green and the negative (over-predictions) in red.

```{r Task 4}
res_t3 <- residuals(model_t3)
plot(ne.df$N ~ ne.df$E, cex=3*abs(res_t3)/max(abs(res_t3)), col=ifelse(res_t3 > 0, "green", "red"), xlab="E", ylab="N",
main="Residuals from 1st-order trend", sub="Positive: green; negative: red", asp=1)
grid()
```

<br>

Q8: What is the spatial pattern of the trend-surface residuals? Does it suggest a higher-order trend?

A: 

- Spatial pattern: large residuals tend to be near each other, and vice-versa. It suggests a higher-order trend.

<br>

# Second-order trend surface
***

Task 5: Repeat all the steps of the previous section, and answer the same questions, for a full 2nd-order trend surface.

```{r Task 5}
model_t5 <- lm(ANN_GDD50 ~ E + N + I(E^2) + I(N^2) +I(E*N), 
               data = ne.df)
```

<br>

Q9: What proportion of the variation in GDD50 is explained by the second-order trend?

A: 

```{r Q9}
summary(model_t5)
```

- The proportion of variability explained is 0.68.

<br>

Q10: Which of the two coördinates (taking into account their squares also) most affects the GDD50?

A: 

- For each km E it increase by -8.336e-05 +5.311309e-05 m (-0.00003024691 m); for each km N it decrease by -1.523e-03 + 8.675252e-05 m (-0.001436247 m). They both are not that important to the predictants. Compared to E, N is kind of more affected to the GDD50.

<br>

Q11: Is there a relation between the fitted values and residuals? If so, what?

A: 

```{r Q11}
par(mfrow = c(1,2))
plot(model_t5, which = 1:2)
```

- Relation of fits vs. residuals seen in 1st order trend has been removed. But systematic over-prediction of highest values.

<br>

Q12: Are the residuals normally-distributed? If not, how so? Are there any especially large residuals?

A: 

- No, they are not normally distributed. There are over-prediction of highest values. There are especially large residuals. 

<br>

Optional Task: Identify the most poorly-modelled station.


Optional Q: Where is this? Why might it be so poorly modelled?

Optional A: 

Task 6: Display a post-plot of the residuals, with the positive residuals (under-predictions) shown in green and the negative (over-predictions) in red.

```{r Task 6}
res_t5 <- residuals(model_t5)
plot(ne.df$N ~ ne.df$E, cex=3*abs(res_t5)/max(abs(res_t5)), col=ifelse(res_t5 > 0, "green", "red"), xlab="E", ylab="N", main="Residuals from 2nd-order trend", sub="Positive: green; negative: red", asp=1)
grid()
```

<br>

Q13: What is the spatial pattern of the trend-surface residuals? Does it suggest a higher-order trend?

A: 

- These residuals form local clusters of positive, negative, and near-zero; there does not appear to be any overall spatial pattern. So, a higher-order trend surface is not indicated.

<br>

# Comparing trend orders
***

Task 7: Compare the two orders of trend surfaces with an analysis of variance.

```{r Task 7}
anova(model_t3, model_t5)
```

<br>

Q14: Is the 2nd-order trend clearly superior to the 1st-order trend?

A: 

- P-values are `***`, suggesting that most independent variables are significant at 0.000 level, so the second-order surface is statistically superior to the first-order surface.

<br>

# Mapping
***

Task 8: Load the `raster` package into the workspace. 

```{r Task 8(1), eval=FALSE}
install.packages("raster")
```

<br>

Then, load a gridded DEM covering the study area `dem_nj_ny_pa_vt_4km.RData`, with the `load` function. Use the `verbose=TRUE` optional argument to display the names of the objects that are loaded.

```{r Task 8(2)}
load("dem_nj_ny_pa_vt_4km.RData", verbose = TRUE)
```

<br>

Q15: What object is loaded? What is its class? What is its CRS? Which of the point datasets does this match?

A: 

- Loading objects: `dem.ne.m.sp`

```{r Q15}
class(dem.ne.m.sp)
proj4string(dem.ne.m.sp)
```

<br>

Q16: What is the bounding box and resolution of the grid? (You can see these with the `summary` function.)

A: 

```{r Q16(1), warning=FALSE}
summary(dem.ne.m.sp)
```

- bounding box

```{r Q16(2)}
range(dem.ne.m.sp$E)
range(dem.ne.m.sp$N)
```

<br>

- The resolution of grid is 3450m * 3704m. 

<br>

Task 9: Display the grid with the `spplot` function.

```{r Task 9, warning=FALSE}
(grid_t9 <- spplot(dem.ne.m.sp, xlab="East", ylab="North"))
```

<br>

Task 10: Predict over this study area with the 2nd-order trend surface. Add this as a field in the study area data frame.

```{r Task 10}
pred_t10 <- predict.lm(model_t5, newdata = as.data.frame(dem.ne.m.sp))
dem.ne.m.sp$PRED_2ndTS <- predict(model_t5, newdata=as.data.frame(dem.ne.m.sp))
```

<br>

Task 11: Summarize the predictions.

```{r Task 11}
summary(pred_t10)
```

<br>

Q17: How do the predictions compare with the summary of the weather stations observations?

```{r Q17}
summary(ne.df$ANN_GDD50)
summary(pred_t10)
```

<br>

A:

Task 12: Display a map of the predictions using the `spplot` function, and specifying the field to display with the `zcol` argument to `spplot`.

```{r Task 12, warning=FALSE}
(grid_t12 <- spplot(dem.ne.m.sp, zcol="ELEVATION_",
                    xlab="East", ylab="North"))
```


Q18: Is this trend surface a realistic representation of the growing-degree days?  Explain.

A: 

Yes it is. The figure shows realistic geological information, such as the curve line indicating ravers or valleies.

<br>

# Generalized Additive Model (GAM)
***

Task 13: Display a scatterplot of the GDD50 against the two coördinates.

```{r Task 13, warning=FALSE}
require(ggplot2)
plote_t13 <- ggplot(ne.df, aes(x=E, y=ANN_GDD50)) + geom_point() +
   geom_smooth(method = "loess")
plotn_t13 <- ggplot(ne.df, aes(x=N, y=ANN_GDD50)) + geom_point() +
   geom_smooth(method = "loess")
require(gridExtra)
grid.arrange(plote_t13, plotn_t13, ncol = 2)
```

<br>

Q19: Do these relations look linear or quadratic throughout their ranges?

A: 

The relation with North seems almost linear. However, the relation with East is much more scattered.

<br>

Task 14: Compute a Generalized Additive Model (GAM) of GDD50 modelled by the two coördinates.

```{r Task 14}
require(mgcv)
model_t14 <- gam(ANN_GDD50 ~ s(E, N), data = ne.df)
```

<br>

Q20: What proportion of the variation in GDD50 is explained by the GAM? How does this compare to the 2nd-order trend surface?

```{r Q20}
summary(model_t14)
```

A: 

- The proportion of variability explained is 0.77.
- Compared with 2nd-order trend surface, the proportion becomes higher. 

<br>

Task 15: Predict over this study area with the fitted GAM. Add this as a field in the study area data frame.

```{r Task 15}
pred_t15 <- predict.gam(model_t14, newdata = as.data.frame(dem.ne.m.sp))
dem.ne.m.sp$PRED_GAM <- pred_t15
```

Note: you will have to convert the `SpatialPixelsDataFrame` to a regular `data.frame` as the argument to `newdata`, because `predict.gam` does not recognize `sp` objects.

<br>

Task 16: Summarize the predictions.

```{r Task 16}
summary(pred_t15)
```

<br>

Q21: How do the predictions compare with the summary of the weather stations observations, and to the summary of the 2nd-order trend surface predictions?

A:

```{r Q21}
summary(ne.df$ANN_GDD50)
summary(pred_t10)
summary(pred_t15)
```

<br>

Task 17: Display a post-plot of the residuals.

```{r Task 17}
res_t17 <- residuals.gam(model_t14)
plot(ne.df$N ~ ne.df$E, cex=3*abs(res_t17)/max(abs(res_t17)), col=ifelse(res_t17 > 0, "green", "red"), xlab="E", ylab="N",
main="Residuals from GAM", sub="Positive: green; negative: red", asp=1)
grid()

```

<br>

Q22: What is the spatial pattern of the GAM trend-surface residuals? How does this differ from the pattern of the 2nd-order trend residuals?

- No residual spatial correlation. Compared with the 2nd-order trend residuals, GAM trend-surface residuals doesn't appear any spatial pattern. 

<br>

A: 

Task 18: Display a map of the predictions using the `spplot` function.

```{r Task 18}
library(sp)
(map_t18 <- spplot(dem.ne.m.sp, zcol="PRED_GAM",
                   xlab="East", ylab="North"))
```

<br>

Q23: Is this trend surface a realistic representation of the growing-degree days? Explain. How does it improve on the 2nd-order trend?

A: 

- Yes, it is. It improved by including elevation data in the prediction. Independent marginal effect of predictors: 2D trend, 1D elevation. Removes spatial dependence of OLS residuals at the range of the empirical smoother.

<br>


# Thin-plate splines
***

Task: Load the `fields` package, to use for thin-plate spline calculation.



Task: Add a field to the stations dataframe, containing a matrix of the two coördinates, as required by the `fields` package.


Task: compute the thin-plate spline for the growing degree days, as a function of coördinates.



Task: Predict with the thin-plate spline over the study area. Note you will have to make a matrix from the grid coördinates, interpolate on that, and then convert the prediction to a field in the grid.


Task: Display the map.


Q: Is this trend surface a realistic representation of the growing-degree days? Explain. Does it improve on the GAM trend?

A: 

# Challenge -- including elevation

Of course, elevation also affects temperature and thus the growing-degree days. 

Repeat the exercise, but including elevation in all the models. Comment on how and why the trend surfaces (now 3D) are different/improved from the 2D trends.

In these models, consider elevation as an independent dimension, with no interaction with the geographic coördinates (although, that may not be a valid assumption... you could test this... ).

Also, see if a transformation of elevation might make its relation with the target variable more linear.

# Saving the workspace 

Task: Save the workspace for a later exercise using the `save.image` function:


